---
title: "Exercise 1: Multinomial logit model"
author: Kenneth Train and Yves Croissant
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exercise 1: The multinomial logit model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r label = setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, widtht = 65)
options(width = 65)
```

@. The problem set uses data on choice of heating system in California
houses. The data set `Heating` from the `mlogit` package
contains the data in `R` format. The observations consist of
single-family houses in California that were newly built and had
central air-conditioning. The choice is among heating systems. Five
types of systems are considered to have been possible:

- gas central (`gc`),
- gas room (`gr`),
- electric central (`ec`),
- electric room (`er`),
- heat pump (`hp`). 

There are 900 observations with the following variables:

- `idcase` gives the observation number (1-900),
- `depvar` identifies the chosen alternative (`gc`, `gr`, `ec`, `er`, `hp`),
- `ic.alt` is the installation cost for the 5 alternatives,
- `oc.alt` is the annual operating cost for the 5 alternatives,
- `income` is the annual income of the household,
- `agehed` is the age of the household head,
- `rooms` is the number of rooms in the house,
- `region` a factor with levels `ncostl` (northern coastal
  region), `scostl` (southern coastal region), `mountn`
  (mountain region), `valley` (central valley region).

Note that the attributes of the alternatives, namely, installation
cost and operating cost, take a different value for each
alternative. Therefore, there are 5 installation costs (one for each
of the 5 systems) and 5 operating costs. To estimate the logit model,
the researcher needs data on the attributes of all the alternatives,
not just the attributes for the chosen alternative. For example, it is
not sufficient for the researcher to determine how much was paid for
the system that was actually installed (ie., the bill for the
installation). The researcher needs to determine how much it would
have cost to install each of the systems if they had been
installed. The importance of costs in the choice process (i.e., the
coefficients of installation and operating costs) is determined
through comparison of the costs of the chosen system with the costs of
the non-chosen systems.
 
For these data, the costs were calculated as the amount the system
would cost if it were installed in the house, given the
characteristics of the house (such as size), the price of gas and
electricity in the house location, and the weather conditions in the
area (which determine the necessary capacity of the system and the
amount it will be run.) These cost are conditional on the house having
central air-conditioning. (That's why the installation cost of gas
central is lower than that for gas room: the central system can use
the air-conditioning ducts that have been installed.)

In a logit model, each variable takes a different value in each
alternative. So, in our case, for example, we want to know the
coefficient of installation cost in the logit model of system
choice. The variable installation cost in the model actually consists
of five variables in the dataset: `ic.gc`, `ic.gr`, `ic.ec`,
`ic.er` and `ic.hp`, for the installation costs of the five
systems. In the current code, there are two variables in the logit
model. The first variable is called `ic` for installation
cost. This variable consists of five variables in the dataset:
`ic.gc` in the first alternative, `ic.gr` in the second
alternative, etc.

2. Run a model with installation cost and operating cost, without intercepts

(a) Do the estimated coefficients have the expected signs?

```{r }
library("mlogit")
data("Heating", package = "mlogit")
H <- mlogit.data(Heating, shape = "wide", choice = "depvar", varying = c(3:12))
m <- mlogit(depvar ~ ic + oc | 0, H)
summary(m)
``` 

> Yes, they are negative as expected, meaning that as the cost of a
> system rises (and the costs of the other systems remain the same) the
> probability of that system being chosen falls.

(b) Are both coefficients significantly different from zero?

> Yes, the t-statistics are greater than 1.96, which is the critical
> level for 95\% confidence level.


(c) How closely do the average probabilities match the shares of
customers choosing each alternative?

```{r }
apply(fitted(m, outcome = FALSE), 2, mean)
```  

> Not very well. 63.67\% of the sample chose `gc` (as shown at the top of
> the summary) and yet the estimated model gives an average probability
> of only 51.695\%. The other alternatives are also fairly poorly
> predicted. We will find how to fix this problem in one of the models
> below.

(d) The ratio of coefficients usually provides economically meaningful
information. The willingness to pay ($wtp$) through higher installation
cost for a one-dollar reduction in operating costs is the ratio of the
operating cost coefficient to the installation cost coefficient. What
is the estimated $wtp$ from this model? Is it reasonable in magnitude?

$$
U = \beta_{ic} ic + \beta_{oc} oc
$$

$$
dU = \beta_{ic} dic + \beta_{oc} doc = 0 \Rightarrow 
-\frac{dic}{doc}\mid_{dU=0}=\frac{\beta_{oc}}{\beta_{ic}}
$$

```{r }
coef(m)["oc"]/coef(m)["ic"]
```   

> The model implies that the decision-maker is willing to pay \$.73 (ie.,
> 73 cents) in higher installation cost in order to reduce annual
> operating costs by \$1.

> A \$1 reduction in annual operating costs recurs each year. It is
> unreasonable to think that the decision-maker is only willing to pay
> only 73 cents as a one-time payment in return for a \$1/year stream of
> saving. This unreasonable implication is another reason (along with
> the inaccurate average probabilities) to believe this model is not so
> good. We will find below how the model can be improved.

(e) We can use the estimated $wtp$ to obtain an estimate of the
discount rate that is implied by the model of choice of operating
system. The present value of the future operating costs is the
discounted sum of operating costs over the life of the system:
$PV=\sum_{t=1}^L\frac{OC}{(1+r)^t}$ where $r$ is the discount rate and
$L$ being the life of the system. As $L$ rises, the $PV$ approaches
$OC/r$. Therefore, for a system with a sufficiently long life (which
we will assume these systems have), a one-dollar reduction in $OC$
reduces the present value of future operating costs by $(1/r)$. This
means that if the person choosing the system were incurring the
installation costs and the operating costs over the life of the
system, and rationally traded-off the two at a discount rate of $r$,
the decisionmaker's $wtp$ for operating cost reductions would be
$(1/r)$. Given this, what value of $r$ is implied by the estimated
$wtp$ that you calculated in part (c)? Is this reasonable?


> $U=a LC$ where $LC$ is lifecycle cost, equal to the sum of
>   installation cost and the present value of operating costs:
>   $LC=IC+(1/r)OC$. Substituting, we have $U=aIC + (a/r)OC$.

> The models estimates $a$ as $-0.00623$ and $a/r$ as $-0.00457$. So $r =
>   a/(a/r)=-.000623/.00457 = 1.36$ or 136\% discount rate.
>   This is not reasonable, because it is far too high.

@. Estimate a model that imposes the constraint that $r=0.12$ (such that
$wtp=8.33$). Test the hypothesis that $r=0.12$.


> To impose this constraint, we create a lifecycle cost that embodies
> the constraint $lcc=ic+oc/0.12$ and estimate the model with this
> variable.

```{r }
H$lcc <- H$ic + H$oc / 0.12
mlcc <- mlogit(depvar ~ lcc | 0, H)
library("lmtest")
lrtest(m, mlcc)
qchisq(0.05, df = 1, lower.tail = FALSE)
```   

> We perform a likelihood ratio test. The $\ln L$ for this constrained model
> is $-1248.7$. The $\ln L$ for the unconstrained model is $-1095.2$. The test
> statistic is twice the difference in $\ln L$: $2(1248.7-1095.2)=307$. This
> test is for one restriction (ie a restiction on the relation of the
> coefficient of operating cost to that of installation cost.) We
> therefore compare $307$ with the critical value of chi-squared with
> $1$ degree of freedom. This critical value for 95\% confidence is
> $3.8$. Since the statistic exceeds the critical value, we reject the
> hypothesis that $r=0.12$.

@. Add alternative-specific constants to the model. With $J$
alternatives, at most $J-1$ alternative-specific constants can be
estimated. The coefficients of $J-1$ constants are interpreted as
relative to alternative $J$th alternative. Normalize the constant for
the alternative `hp` to 0.

(a) How well do the estimated probabilities match the shares of
customers choosing each alternative? 

```{r }
mc <- mlogit(depvar ~ ic + oc, H, reflevel = 'hp')
summary(mc)
apply(fitted(mc, outcome = FALSE), 2, mean)
```   

> Note that they match exactly: alternative-specific constants in a
> logit model insure that the average probabilities equal the observed
> shares.

(b) Calculate the $wtp$ and discount rate $r$ that is implied by the
estimates. Are these reasonable?

```{r }
wtp <- coef(mc)["oc"] / coef(mc)["ic"]
wtp
r <- 1 / wtp
r
``` 

> The decision-maker is willing to pay \$4.56 for a \$1 year stream of
> savings. This implies $r = 0.22$. The decision-maker applies a 22\%
> discount rate.  These results are certainly more reasonable that in
> the previous model. The decision-maker is still estimated to be
> valuing saving somewhat less than would seem rational (ie applying a
> higher discount rate than seems reasonable). However, we need to
> remember that the decision-maker here is the builder. If home buyers
> were perfectly informed, then the builder would adopt the buyer's
> discount rate. However, the builder would adopt a higher discount rate
> if home buyers were not perfectly informed about (or believed) the
> stream of saving.

(c) This model contains constants for all alternatives
`ec`-`er`-`gc`-`gr`, with the constant for alternative
`hp` normalized to zero. Suppose you had included constants for
alternatives `ec`-`er`-`gc`-`hp`, with the constant for
alternative `gr` normalized to zero. What would be the estimated
coefficient of the constant for alternative `gc`? Figure this out
logically rather than actually estimating the model.

> We know that when the hp is left out, the constant for alternative
> `gc` is $1.71074$ meaning that the average impact of unicluded factors
> is $1.71074$ higher for alternative `gc` than for alternative
> hp. Similarly, the constant for alternative `gr` is $0.30777$. If `gr`
> were left out instead of `hp`, then all the constants would be
> relative to alternative `gr`. The constant for alternative `gc` would
> the be $1.71074-.30777=1.40297$. That is, the average impact of
> unincluded factors is $1.40297$ higher for alt `gc` than alt
> `gr`. Similarly for the other alternatives.  Note the the constant for
> alt 5 would be $0-.30777=-.3077$, since `hp` is normalized to zero in
> the model with `hp` left out.

```{r }
update(mc, reflevel = "gr")
``` 

@. Now try some models with sociodemographic variables entering.

(a) Enter installation cost divided by income, instead of installation
cost. With this specification, the magnitude of the installation cost
coefficient is inversely related to income, such that high income
households are less concerned with installation costs than lower
income households. Does dividing installation cost by income seem to
make the model better or worse?

```{r }
mi <- mlogit(depvar ~ oc + I(ic / income), H)
summary(mi)
```   

> The model seems to get worse. The $\ln L$ is lower (more negative) and the
> coefficient on installation cost becomes insignificant (t-stat below
> 2).


(b) Instead of dividing installation cost by income, enter
alternative-specific income effects.  What do the estimates imply
about the impact of income on the choice of central systems versus
room system? Do these income terms enter significantly?

```{r }
mi2 <- mlogit(depvar ~ oc + ic | income, H, reflevel = "hp")
```   

> The model implies that as income rises, the probability of heat pump
> rises relative to all the others (since income in the heat pump alt is
> normalized to zero, and the others enter with negative signs such that
> they are lower than that for heat pumps. Also, as income rises, the
> probability of gas room drops relative to the other non-heat-pump
> systems (since it is most negative).

> Do these income terms enter significantly? No. It seems that income
> doesn't really have an effect. Maybe this is because income is for the
> family that lives in the house, whereas the builder made decision of
> which system to install.

```{r }
lrtest(mc, mi2)
waldtest(mc, mi2)
scoretest(mc, mi2)
``` 

(c) Try other models. Determine which model you think is best from these data.

> I'm not going to give what I consider my best model: your ideas on
> what's best are what matter here.


6. We now are going to consider the use of logit model for
prediction. Estimate a model with installation costs, operating costs,
and alternative specific constants. Calculate the probabilities for
each house explicitly. Check to be sure that the mean probabilities
are the same as you got in exercise 4.

```{r }
X <- model.matrix(mc)
alt <- index(H)$alt
chid <- index(H)$chid
eXb <- as.numeric(exp(X %*% coef(mc)))
SeXb <- tapply(eXb, chid, sum)
P <- eXb / SeXb[chid]
P <- matrix(P, ncol = 5, byrow = TRUE)
head(P)
apply(P, 2, mean)
```

> This can be computed much more simply using the \Rf{fitted` function,
> with the \Ra{outcome`{fitted` argument set to \Rv{FALSE` so that the
> probabilities for all the alternatives (and not only the chosen one)
> is returned.

```{r }
apply(fitted(mc, outcome = FALSE), 2, mean)
``` 

@. The California Energy Commission (CEC) is considering whether to
offer rebates on heat pumps. The CEC wants to predict the effect of
the rebates on the heating system choices of customers in
California. The rebates will be set at 10\% of the installation
cost. Using the estimated coefficients from the model in exercise 6,
calculate new probabilities and predicted shares using the new
installation cost of heat pump. How much do the rebates raise the
share of houses with heat pumps?  

```{r }
Hn <- H
Hn[Hn$alt == "hp", "ic"] <- 0.9 * Hn[Hn$alt == "hp", "ic"]
apply(predict(mc, newdata = Hn), 2, mean)
```

> We estimate the model with the actual costs. Then we change the costs
> and calculate probabilities with the new costs. The average
> probability is the predicted share for an alternative. At the original
> costs, the heat pump share is $0.0555$ (ie, about 5.5\%) This share is
> predicted to rise to $0.0645$ (about 6.5\%) when rebates are given.

@. Suppose a new technology is developed that provides more efficient
central heating. The new technology costs \$200 more than the central
electric system. However, it saves 25\% of the electricity, such that
its operating costs are 75\% of the operating costs of `ec`. We want
to predict the potential market penetration of this technology. Note
that there are now six alternatives: the original five alternatives
plus this new one. Calculate the probability and predict the market
share (i.e., the average probability) for all six alternatives, using
the model that is estimated on the original five alternatives. (Be
sure to use the original installation cost for heat pumps, rather than
the reduced cost in exercise 7.) What is the predicted market share
for the new technology? From which of the original five systems does
the new technology draw the most customers? 

```{r }
X <- model.matrix(mc)
Xn <- X[alt == "ec",]
Xn[, "ic"] <- Xn[, "ic"] + 200
Xn[, "oc"] <- Xn[, "oc"] * 0.75
unchid <- unique(index(H)$chid)
rownames(Xn) <- paste(unchid, 'new', sep = ".")
chidb <- c(chid, unchid)
X <- rbind(X, Xn)
X <- X[order(chidb), ]
eXb <- as.numeric(exp(X %*% coef(mc)))
SeXb <- as.numeric(tapply(eXb, sort(chidb), sum))
P <- eXb / SeXb[sort(chidb)]
P <- matrix(P, ncol = 6, byrow = TRUE)
apply(P, 2, mean)
```   

> The new technology captures a market share of 0.1036. That is, it gets
> slightly more than ten percent of the market.

> It draws the same percent (about 10\%) from each system. This means
> that it draws the most in absolute terms from the most popular system,
> gas central. For example, gas central drops from to $0.637$ to
> $0.571$; this is an absolute drop of $0.637-0.571=0.065$ and a percent
> drop of $0.065/0.637$ about 10\%. Of the 10.36\% market share that is
> attained by the new technology, 6.5\% of it comes from gas
> central. The other systems drop by about the same percent, which is
> less in absolute terms.

> The same percent drop for all systems is a consequence of the IIA
> property of logit. To me, this property seems unreasonable in this
> application. The new technology is a type of electric system. It seems
> reasonable that it would draw more from other electric systems than
> from gas systems. Models like nested logit, probit, and mixed logit
> allow more flexible, and in this case, more realistic substitution
> patterns.


