---
title: "Logit models relaxing the iid hypothesis"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{4. Logit models relaxing the iid hypothesis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r label = setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, widtht = 65)
options(width = 65)
```

In the previous section, we assumed that the error terms were iid
(identically and independently distributed), i.e., uncorrelated and
homoscedastic. Extensions of the basic multinomial logit model have
been proposed by relaxing one of these two hypothesis while
maintaining the hypothesis of a Gumbel distribution.

## The heteroskedastic logit model

The heteroskedastic logit model was proposed by @BHAT:95.  The
probability that $U_l>U_j$ is:

$$
P(\epsilon_j<V_l-V_j+\epsilon_l)=e^{-e^{-\frac{(V_l-V_j+\epsilon_l)}{\theta_j}}},
$$

which implies the following conditional and unconditional
probabilities

\begin{equation*}
  (P_l \mid \epsilon_l) =\prod_{j\neq
    l}e^{-e^{-\frac{(V_l-V_j+\epsilon_l)}{\theta_j}}},
\end{equation*}

\begin{equation*}
  \begin{array}{rcl}
  P_l&=&\displaystyle\int_{-\infty}^{+\infty} \prod_{j\neq l}
  \left(e^{-e^{-\frac{(V_l-V_j+t)}{\theta_j}}}\right)\frac{1}{\theta_l}e^{-\frac{t}{\theta_l}}e^{-e^{-\frac{t}{\theta_l}}}
  dt\\
 &=& \displaystyle \int_{0}^{+\infty}\left(e^{-\sum_{j \neq
      l}e^{-\frac{V_l-V_j-\theta_l \ln t}{\theta_j}}}\right)e^{-t}dt.
     \end{array}
\end{equation*}

There is no closed form for this integral, but it can be efficiently
computed using a Gauss quadrature method, and more precisely the
Gauss-Laguerre quadrature method.

## The nested logit model

The nested logit model was first proposed by [@MCFAD:78]. It is a
generalization of the multinomial logit model that is based on the
idea that some alternatives may be joined in several groups (called
nests). The error terms may then present some correlation in the same
nest, whereas error terms of different nests are still uncorrelated.

Denoting $m=1... M$ the nests and $B_m$ the set of alternatives
belonging to nest $m$, the cumulative distribution of the errors is:

$$
\mbox{exp}\left(-\sum_{m=1}^M \left( \sum_{j \in B_m}
    e^{-\epsilon_j/\lambda_m}\right)^{\lambda_m}\right).
$$

The marginal distributions of the $\epsilon$s are still univariate
extreme value, but there is now some correlation within
nests. $1-\lambda_m$ is a measure of the correlation, i.e., $\lambda_m
= 1$ implies no correlation. In the special case where $\lambda_m=1\;
\forall m$, the errors are iid Gumbel errors and the nested logit
model reduce to the multinomial logit model.  It can then be shown
that the probability of choosing alternative $j$ that belongs to nest
$l$ is:

$$
P_j = \frac{e^{V_j/\lambda_l}\left(\sum_{k \in B_l}
    e^{V_k/\lambda_l}\right)^{\lambda_l-1}} {\sum_{m=1}^M\left(\sum_{k
      \in B_m} e^{V_k/\lambda_m}\right)^{\lambda_m}},
$$

and that this model is a random utility model if all the $\lambda$
parameters are in the $0-1$ interval.^[A slightly different
version of the nested logit model [@DALY:87] is often used, but is not
compatible with the random utility maximization hypothesis. Its
difference with the previous expression is that the deterministic
parts of the utility for each alternative is not divided by the nest
elasticity. The differences between the two versions have been
discussed in @KOPP:WEN:98, @HEIS:02 and @HENS:GREEN:02.]

Let us now write the deterministic part of the utility of alternative
$j$ as the sum of two terms: the first one ($Z_j$) being specific to
the alternative and the second one ($W_l$) to the nest it belongs to:

$$V_j=Z_j+W_l.$$

We can then rewrite the probabilities as follow:

$$
\begin{array}{rcl}  
P_j&=&\frac{e^{(Z_j+W_l)/\lambda_l}}{\sum_{k \in B_l}
  e^{(Z_k+W_l)/\lambda_l}}\times \frac{\left(\sum_{k \in B_l}
    e^{(Z_k+W_l)/\lambda_l}\right)^{\lambda_l}}
{\sum_{m=1}^M\left(\sum_{k \in B_m}
    e^{(Z_k+W_m)/\lambda_m}\right)^{\lambda_m}}\\
&=&\frac{e^{Z_j/\lambda_l}}{\sum_{k \in B_l}
    e^{Z_k/\lambda_l}}\times 
\frac{\left(e^{W_l/\lambda_l}\sum_{k \in B_l} e^{Z_k/\lambda_l}\right)^{\lambda_l}}
{\sum_{m=1}^M\left(e^{W_m/\lambda_m}\sum_{k
      \in B_m} e^{Z_k/\lambda_m}\right)^{\lambda_m}}.
\end{array}
$$

Then denote $I_l=\ln \sum_{k \in B_l} e^{Z_k/\lambda_l}$ which is
often called the log-sum, the inclusive value or the inclusive
utility.^[We've already encountered this expression in
Section~\@ref(sec:conssurplus).] We then can write the probability of
choosing alternative $j$ as:

$$
P_j=\frac{e^{Z_j/\lambda_l}}{\sum_{k \in B_l}
    e^{Z_k/\lambda_l}}\times 
\frac{e^{W_l+\lambda_l I_l}}{\sum_{m=1}^Me^{W_m+\lambda_m I_m}}.
$$

The first term $\mbox{P}_{j\mid l}$ is the conditional probability of
choosing alternative $j$ if nest $l$ is chosen. It is often referred
as the *lower model*. The second term $\mbox{P}_l$ is the marginal
probability of choosing nest $l$ and is referred as the *upper model*.
$W_l+\lambda_l I_l$ can be interpreted as the expected utility of
choosing the best alternative in $l$, $W_l$ being the expected utility
of choosing an alternative in this nest (whatever this alternative is)
and $\lambda_l I_l$ being the expected extra utility gained by being
able to choose the best alternative in the nest.  The inclusive values
link the two models.  It is then straightforward to show that IIA
applies within nests, but not for two alternatives in different nests.

A consistent but inefficient way of estimating the nested logit model
is to estimate separately its two components. The coefficients of the
lower model are first estimated, which enables the computation of the
inclusive values $I_l$. The coefficients of the upper model are then
estimated, using $I_l$ as covariates. Maximizing directly the
likelihood function of the nested model leads to a more efficient
estimator.

## Applications

### ModeCanada

@BHAT:95 estimated the heteroscedastic logit model on the
`ModeCanada` data set. Using `mlogit`, the heteroscedastic
logit model is obtained by setting the `heterosc` argument
to `TRUE`:

```{r label = 'heteroscedastic model for the ModeCanada data'}
library("mlogit")
data("ModeCanada", package = "mlogit")
MC <- mlogit.data(ModeCanada, subset = noalt == 4, chid.var = "case",
                  alt.var = "alt", drop.index = TRUE)
ml.MC <- mlogit(choice ~ freq + cost + ivt + ovt | urban + income, MC, 
reflevel = 'car', alt.subset = c("car", "train", "air"))
hl.MC <- mlogit(choice ~ freq + cost + ivt + ovt | urban + income, MC, 
reflevel = 'car', alt.subset = c("car", "train", "air"), heterosc = TRUE)
coef(summary(hl.MC))[11:12, ]
``` 

The variance of the error terms of train and air are respectively
higher and lower than the variance of the error term of car (set to
1). Note that the z-values and p-values of the output are not
particularly meaningful, as the hypothesis that the coefficient is
zero (and not one) is tested.  The homoscedasticity hypothesis can be
tested using any of the three tests. A particular convenient syntax is
provided in this case.  For the likelihood ratio and the wald test,
one can use only the fitted heteroscedastic model as argument. In this
case, it is guessed that the hypothesis that the user wants to test is
the homoscedasticity hypothesis.

```{r label = 'homoscedasticity tests: lr and Wald (1)'}
lr.heter <- lrtest(hl.MC, ml.MC)
wd.heter <- waldtest(hl.MC, heterosc = FALSE)
``` 

or, more simply:

```{r label = 'homoscedasticity tests: lr and Wald (2)', results = 'hide'}
lrtest(hl.MC)
waldtest(hl.MC)
``` 

The Wald test can also be computed using the `linearHypothesis`
function from the `car` package:

```{r label = 'homoscedasticity tests: Wald test'}
library("car")
lh.heter <- linearHypothesis(hl.MC, c('sp.air = 1', 'sp.train = 1'))
``` 

For the score test, we provide the constrained model as argument,
which is the standard multinomial logit model and the supplementary
argument which defines the unconstrained model, which is in this case
`heterosc = TRUE`.

```{r label = 'homoscedasticity tests: score test'}
sc.heter <- scoretest(ml.MC, heterosc = TRUE)
``` 

```{r label = 'convenient statpval function 2', include = FALSE}
statpval <- function(x){
    if (inherits(x, "anova")) 
        result <- as.matrix(x)[2, c("Chisq", "Pr(>Chisq)")]
    if (inherits(x, "htest")) result <- c(x$statistic, x$p.value)
    names(result) <- c("stat", "p-value")
    round(result, 3)
}
``` 


```{r label = 'homoscedasticity tests: results'}
sapply(list(wald = wd.heter, lh = lh.heter, score = sc.heter,
lr = lr.heter), statpval)
``` 

The homoscedasticity hypothesis is strongly rejected using the Wald
test, but only at the 1 and 5\% level for, respectively, the score and
the likelihood ratio tests.

### JapaneseFDI

@HEAD:MAYE:04 analyzed the choice of one of the 57 European
regions belonging to 9 countries by Japenese firms to implement a new
production unit.

```{r label = 'loading the JapaneseFDI data set'}
data("JapaneseFDI", package = "mlogit")
jfdi <- mlogit.data(JapaneseFDI, chid.var = "firm", alt.var = "region",
group.var = "country")
``` 

Note that we've used an extra argument to `mlogit.data` called
`group.var` which indicates the grouping variable,
which will be used later to define easily the nests.  There are two
sets of covariates: the wage rate `wage`, the unemployment rate
`unemp`, a dummy indicating that the region is eligible to European
funds `elig` and the area `area` are observed at the regional
level and are therefore relevant for the estimation of the lower
model, whereas the social cotisation rate `scrate` and the
corporate tax rate `ctaxrate` are observed at the country level and
are therefore suitable for the upper model.

We first estimate a multinomial logit model:
```{r label = 'multinomial logit for JapaneseFDI'} 
ml.fdi <- mlogit(choice ~ log(wage) + unemp + elig + log(area) + scrate +
ctaxrate | 0, data = jfdi)
``` 

Note that, as the covariates are only alternative specific, the
intercepts are not identified and therefore have been removed.  We
next estimate the lower model, which analyses the choice of a region
within a given country. Therefore, for each choice situation, we
estimate the choice of a region on the subset of regions of the
country which has been chosen. Moreover, observations concerning
Portugal and Ireland are removed as these two countries are
mono-region.

```{r label = 'lower model estimation'}
lm.fdi <- mlogit(choice ~ log(wage) + unemp + elig + log(area) | 0,
data = jfdi, subset = country == choice.c & ! country %in% c("PT", "IE"))
``` 

We next use the fitted lower model in order to compute the inclusive
value, at the country level:

$$
\mbox{iv}_{ig} = \ln \sum_{j \in B_g} e^{\beta^\top x_{ij}},
$$

where $B_g$ is the set of regions for country $g$. When a grouping
variable is provided in the `mlogit.data` function, inclusive values
are by default computed for every group $g$ (global inclusive values
are obtained by setting the `type` argument to `"global"`). By
default, `output` is set to `"chid"` and the results is a vector (if
`type = "global"`) or a matrix (if `type = "region"`) with row number
equal to the number of choice situations. If `output` is set to
`"obs"`, a vector of length equal to the number of lines of the data
in long format is returned. The following code indicates different
ways to use the `logsum` function:

```{r label = 'use of the logsum function'}
lmformula <- formula(lm.fdi)
head(logsum(ml.fdi, data = jfdi, formula = lmformula, type = "group"), 2)
head(logsum(ml.fdi, data = jfdi, formula = lmformula, type = "global"))
head(logsum(ml.fdi, data = jfdi, formula = lmformula, output = "obs"))
head(logsum(ml.fdi, data = jfdi, formula = lmformula, type = "global",
output = "obs"))
``` 

To add the inclusive values in the original `data.frame`, we use
`output = "obs"` and the `type` argument can be omitted as its default
value is `"group"`:

```{r label = 'adding the logsum to the data'}
JapaneseFDI$iv <- logsum(lm.fdi, data = jfdi, formula = lmformula,
output = "obs")
``` 

We next select the relevant variables for the estimation of the upper
model, select unique lines in order to keep only one observation for
every choice situation / country combination and finally we coerce the
response (`choice.c`) to a logical for the chosen country.

```{r label = 'data suitable for the upper model'}
JapaneseFDI.c <- subset(JapaneseFDI, 
select = c("firm", "country", "choice.c", "scrate", "ctaxrate", "iv"))
JapaneseFDI.c <- unique(JapaneseFDI.c)
JapaneseFDI.c$choice.c <- with(JapaneseFDI.c, choice.c == country)
```

Finally, we estimate the upper model, using the previously computed
inclusive value as a covariate.

```{r label = 'estimation of the upper model'}
jfdi.c <- mlogit.data(JapaneseFDI.c, choice = "choice.c",
alt.var = "country", chid.var = "firm", shape = "long")
um.fdi <- mlogit(choice.c ~ scrate + ctaxrate + iv | 0, data = jfdi.c)
```

If one wants to obtain different `iv` coefficients for different
countries, the `iv` covariate should be introduced in the 3th part
of the formula and the coefficients for the two mono-region countries
(Ireland and Portugal) should be set to 1, using the
`constPar` argument.

```{r label = 'upper model with different iv coefficients'}
um2.fdi <- mlogit(choice.c ~ scrate + ctaxrate | 0 | iv, data = jfdi.c, 
constPar = c("iv:PT" = 1, "iv:IE" = 1))
``` 

We next estimate the full-information maximum likelihood nested
model. It is obtained by adding a `nests` argument to the
`mlogit` function. This should be a named list of alternatives
(here regions), the names being the nests (here the countries). More
simply, if a group variable has been indicated while using
`mlogit.data`, `nests` can be a boolean.

Two flavors of nested models can be estimated, using the `un.nest.el`
argument which is a boolean. If `TRUE`, one imposes that the
coefficient associated with the inclusive utility is the same for
every nest, which means that the degree of correlation inside each
nest is the same. If `FALSE`, a different coefficient is estimated for
every nest.

```{r label = 'nested logit models'}
nl.fdi <- mlogit(choice ~ log(wage) + unemp + elig + log(area) + scrate +
ctaxrate | 0, data = jfdi, nests = TRUE, un.nest.el = TRUE)
nl2.fdi <- update(nl.fdi, un.nest.el = FALSE, constPar = c('iv:PT' = 1,
'iv:IE' = 1))
``` 

The results of the fitted models are presented in the following table:

```{r label = 'results for the JapaneseFDI data', results = 'asis'}
library("texreg")
htmlreg(list('Mult. logit' = ml.fdi, 'Lower model' = lm.fdi,
             'Upper model' = um.fdi, 'Upper model' = um2.fdi, 'Nested logit' = nl.fdi,
             'Nested logit' = nl2.fdi),
        fontsize = "footnotesize", float.pos = "hbt",  label = "tab:nlogit",
        caption = "Choice by Japanese firms of a european region.")
``` 

For the nested logit models, two tests are of particular interest:


- the test of no nests, which means that all the nest elasticities
  are equal to 1,
- the test of unique nest elasticities, which means that all the
  nest elasticities are equal to each other.



For the test of no nests, the nested model is provided as the unique
argument for the `lrtest` and the `waldtest` function. For the
`scoretest`, the constrained model (i.e., the multinomial logit model)
is provided as the first argument and the second argument is `nests`,
which describes the nesting structure that one wants to test.

```{r label = 'test of no nests'}
lr.nest <- lrtest(nl2.fdi)
wd.nest <- waldtest(nl2.fdi)
sc.nest <- scoretest(ml.fdi, nests = TRUE, constPar = c('iv:PT' = 1,
'iv:IE' = 1))
``` 

The Wald test can also be performed using the `linearHypothesis`
function:
```{r label = 'test of no nests with linhyp'}
lh.nest <- linearHypothesis(nl2.fdi, c("iv:BE = 1", "iv:DE = 1",
"iv:ES = 1", "iv:FR = 1", "iv:IT = 1", "iv:NL = 1", "iv:UK = 1"))
``` 

```{r label = 'results of the tests of no nests'}
sapply(list(wald = wd.nest, lh = lh.nest, score = sc.nest, lr = lr.nest),
statpval)
``` 

The three tests reject the null hypothesis of no correlation. We next
test the hypothesis that all the nest elasticities are equal.

```{r label = 'computing the test for equal iv coefficients'}
lr.unest <- lrtest(nl2.fdi, nl.fdi)
wd.unest <- waldtest(nl2.fdi, un.nest.el = TRUE)
sc.unest <- scoretest(ml.fdi, nests = TRUE, un.nest.el = FALSE,
constPar = c('iv:IE' = 1, 'iv:PT' = 1))
lh.unest <- linearHypothesis(nl2.fdi, c("iv:BE = iv:DE", "iv:BE = iv:ES", 
"iv:BE = iv:FR", "iv:BE = iv:IT", "iv:BE = iv:NL", "iv:BE = iv:UK"))
``` 

```{r label = 'results of the tests of equal iv coefficients'}
sapply(list(wald = wd.unest, lh = lh.unest, score = sc.unest,
lr = lr.unest), statpval)
```

Once again, the three tests strongly reject the hypothesis.

## Bibliography

